Improved state-space models for inference about spatial and temporal variation in abundance from count data

### 2013-11-12 Dail_Mad_ext.tex
more edits to intro

\documentclass[12pt]{article}
\renewcommand*\abstractname{Summary}
\usepackage[a4paper,hmargin={2.4cm,2.4cm},vmargin={2.4cm,2.4cm}]{geometry}
\usepackage[mathlines]{lineno} %,displaymath
\usepackage{setspace}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{bm}
\usepackage{graphicx}
\usepackage[round,colon,authoryear]{natbib}
\usepackage{color, soul}
\usepackage{fancyvrb}
\usepackage{verbatim}
\usepackage[pdftex,colorlinks=true,citecolor=black,linkcolor=black,urlcolor=black,pdftex,pdfstartview=FitH,bookmarks=true]{hyperref}
\usepackage{rotating}

@@

\abstract{
\noindent
\begin{enumerate}
\item Models of population dynamics %play a central role in theoretical and
%applied ecology, where they 
are frequently used for purposes such as testing 
hypotheses about density dependence and predicting species' responses 
to future environmental change or conservation actions. Fitting models
of population dynamics to field data is challenging because most data
sets are characterized by missing data and observation error, which
can inflate estimates of process variation if ignored.  
%Failure to account for observation error in such models can result in 
%bias. Therefore, ecologists have increasingly relied on 
Recently, state-space models have been developed to deal with these problems
%these probdirectly model 
by directly modeling both the observation error and the ecological
process of interest.
\item Conventional state-space models, however, have
three important limitations: (1) the parameters are not identifiable
in many common situations, (2) they do not admit spatial variation in
population dynamics, and (3) they assert an implausible model for
%we have no clear interpretation of the
observation error. We demonstrate how each of these problems can be
resolved using a class of hierarchical models proposed by
\citet[Biometrics]{dail_madsen:2011} for spatially-replicated
time-series data that attributes observation error to imperfect detection.
\item We expand this class of models to accommodate classical growth
models (e.g. exponential and Ricker-logistic), zero-inflation,
and random effects such as observer-specific
detection probabilities. We also present methods for forecasting
population size under future environmental conditions. Implementation
of these ideas is possible using either frequentist or Bayesian
methods, as demonstrated by accompanying {\bf R} and {\bf JAGS} code.%. We provide code to fit these models using the \textbf{R} package
%\texttt{unmarked} and \textbf{JAGS} is also provided.
\item Results of a simulation study suggest that bias is %was 
negligible and coverage
nominal for the proposed model extensions. An analysis of data from
the North American Breeding Bird Survey highlights how these methods
can be readily applied to existing data, but it also suggests 
that precision will be low when direct information about detection
probability (such as is collected using distance sampling or
replicated counts) is lacking.
\end{enumerate}
}

@@

Recently developed state-space models %are a widely used approach for
have made it possible to %become the primary tool for 
study population dynamics while accounting for both process variation
and observation error \citep{devalpine_hastings:2002,
  buckland_etal:2004, dennis_etal:2006}. Classical state-space
models are time-series models in which the true state of the
system (e.g. population size during each year) is modeled as a hidden
process, and the observed data are modeled conditional on this process
and the observation error. One reason for the widespread adoption of
state-space models in ecology is that failure to account for 
%process variation and 
observation error can bias estimators of abundance and
population growth parameters. For instance, the strength of
density dependence will be over-estimated if observation error is
ignored \citep{link_nichols:1994,shenk_etal:1998}.

A simple state-space model can be described as follows.
Let $N_t$ be the abundance of a species during year $t$, for
$t=1,\hdots,T$, and let $X_t$ be
the observed data, which differs from $N_t$ due to observation error,
a random effect denoted $\zeta_t$. Temporal variation in $N_t$ is
modeled using a population growth model, $\mu(N_{t-1})$
and random process variation denoted $\eta_t$.
% Richard, is switching this from $\grave{o}$ to $\eta$ okay?  Andy thought
% the o with accent was weird and I think I agree.  $\nu$ seems standard,
% but we use it for a slightly different meaning (ES only) later.  
The growth model may be density-dependent, as in the case of the 
Ricker-logistic model, or it might be density-independent, such as when growth
is exponential.
%, in which $\mu(N_{t-1}) = N_{t-1}e^{r}$ where $r$ is the intrinsic rate of increase.
A generic model can be written: %The full model can now be written as:
\begin{subequations}
  \label{eq:ss1}
  \begin{align}
    N_1 &= X_1 \label{eq:ss1a} \\
N_t &= \mu(N_{t-1}) + \eta_{t-1} \quad \text{for} \;
t=2,\hdots,T \label{eq:ss1b} \\
X_t &= N_t + \zeta_t \qquad \qquad \;\, \text{for} \;
t=1,\hdots,T. \label{eq:ss1c}
  \end{align}
\end{subequations}
%where $\eta_t$ is the random effect allowing for process
%variation unaccounted for by the deterministic model. 
In classical
state-space models, the two sets of random effects
are assumed to be i.i.d. Gaussian deviates:
$\eta_t \sim \mathrm{N}(0, \sigma)$ and
$\zeta_t \sim \mathrm{N}(0, \tau)$. Standard practice is to
ignore process variation associated with $N_1$, as
indicated by Eq~\ref{eq:ss1a}. Alternatively, the population is
assumed to be at equilibrium such that the $N_1$ can be regarded as an
outcome of the equilibrium distribution.

Even though state-space models such as that shown in Eq.~\ref{eq:ss1}
are among the most widespread approaches for modeling population dynamics
using time-series data, several problems exist. First,
time-series data and the underlying abundance parameters of interest
are typically integer-valued, as in the case of count data, raising
concerns about the Gaussian assumption for the random
effects. Second, the model for observation error is implausible. %has little biological basis. 
Third, the model does not account for spatial variation in abundance. Lastly, some of the
parameters of the model are not estimable. We briefly discuss each of
these points before describing a general approach to resolving these
shortcomings.

%Use of the Gaussian distribution for modeling random process variation
%and observation error is often motivated by convenience rather than
%biology. Specifically, t
The assumption that random effects are Gaussian distributed makes it
easier to estimate parameters because methods such as the Kalman
filter can be applied %The Gaussian assumption allows for efficient methods of parameter
%estimation, such as the Kalman filter
\citep{dennis_etal:2006}, which are not available otherwise. % which is much more computationally efficient
%than estimation methods when random effects are not Gaussian distributed
\citep{devalpine_hastings:2002}. %The problem with this is that it
However, these models effectively assume that abundance can be take on any
real value, including negative values, 
%This approach allows for negative and non-integer values of the state-variable, which is
%inconsistent with the observed data, and may 
and these models can result in implausible predictions.
Transforming the count data and the underlying abundance variables 
do little to resolve these issues \citep{ohara_kotze:2010}.

Another problem with standard state-space models is that the model for
the mean zero Gaussian distribution for the observation error
implies that false positive detections are
equally likely as false negatives. To our knowledge, all studies to
date have shown that false negative detections are much more common
that false positives (CITE DAVE MILLER PAPERS). 
%will be
%higher than $N_t$ as often as it is lower than $N_t$. 
%Identifying a biological mechanism that would cause such symmetric errors
%is difficult.
A more likely form of observation error, and one that has been recognized for well
over a century, results from failing to detect individuals that are
present. Imperfect detection may be attributable to
characteristics of the species under study, such as its elusiveness,
or to individuals being missed by the ecologist collecting the data in the field.
Although a vast number of methods have been devised for accounting for
this form of observation error, rarely have these methods been
integrated into state-space models \citep[but
see][]{buckland_etal:2004}.

Common formulations of state space models do not admit
spatial variation. This reduces the scope of the inferences that can
be drawn from the models. Moreover, space is an important factor in the
regulation of natural populations \citep{murdoch:1994,
pulliam:1988, gill_etal:2001}.

A more serious problem %than the ones associated with the Gaussian assumptions 
is that the parameters of simple state-space models
such as Eq~\ref{eq:ss1} are not identifiable in many
circumstances \citep{polansky_etal:2009}. Specifically, abundance in
the first year cannot be estimated because there is only one data
point (the count in the first year) that carries any information about
it. 
%Recall that the
%Eq.~\ref{eq:ss1} did not specify distributions for $N_1$, % or $X_1$;
%although $N_1$ is a random variable. Thus, uncertainty exists
%that should be accounted for in the model.  
%%are needed to fully specify the model.
%To be more specific, a fully-specified state-space model requires at least
%three probability distributions, which we represent
%using bracket notation (i.e., $[Y|\Omega]$ is interpreted
%as the probability distribution of the random variable $Y$ given the
%parameter $\Omega$). The three probability distributions required for a
%state-space model are:
%%%% JEFF, does this extra sentence help? This notation comes from
%%%% several of the original state-space papers and is widespread in
%%%% books like Andy's and Bill's.
%\hl{[I'm not familiar with bracket notation,
% so these equations are confusing to me. How widely used is this
% notation in ecology? Will others likely understand this? Also, we might want
%to save space by trimming a bunch of this discussion back down.]}
%\begin{subequations}
%  \label{eq:ss2}
%  \begin{align}
%    [N_1&|\bm{\theta}] \label{eq:ss2a} \\ \label{eq:ss2b}
%[N_t&|N_{t-1},\bm{\Theta}] \quad \text{for} \; t=2,\hdots,T \\
%\label{eq:ss2c}
%    [X_t&|N_t,\bm{p}]  \qquad \; \text{for} \; t=1,\hdots,T
%  \end{align}
%\end{subequations}
%where $\bm{\theta}$ are the process variation parameters for the
%initial state, $\bm{\Theta}$ are the process variation parameters
%influencing how abundance changes over time, and $\bm{p}$ are the
%observation error parameters. As mentioned previously, Eq.~\ref{eq:ss2b}
%and Eq.~\ref{eq:ss2c} are assumed to be Gaussian in classical state-space
%models, but what should be the distribution for Eq.~\ref{eq:ss2a}?
%The three lines of general model
%described by Eq.~\ref{eq:ss2} correspond to the three equations shown
%for the specific example in Eq~\ref{eq:ss1}. However, in this case,
%we allow $N_1$ to be a random variable. But what should its
%distribution, $[N_1|\bm{\theta}]$ be?
%And how can $\bm{\theta}$ be estimated given that there is only a single observation
%vailable? 
%Often, these problems are simply assumed away. For
%example, sometimes researchers assume that there is no process variation in
%the first year, although this seems unjustified. Alternatively, the
%population may be assumed to be at equilibrium such that the expected
%value of $N_t$ is constant through time. Although this 
%makes the parameters $\bm{\theta}$ identifiable, assuming equilibrium
%defeats the objective of many studies of population dynamics, namely
%determining why a population varies over time.
%% OR
%Sometimes researchers assume that there is no process variation in 
%the first year or that the population is at equilibrium such that the expected
%value of $N_t$ is constant through time, but these approaches are unsatisfactory.

@@

In this paper, we focus on the model of \citet[henceforth the DM model]{dail_madsen:2011}
that resolves many of the problems with classical state-space models.
%simultaneous resolves each of the aforementioned problems with
%traditional state-space models, and is designed for simple count data.
%In the following section, we describe the DM model in its original
%form and explain how it resolves each of the deficiencies with
%standard state-space models. 
In Section~\ref{sec:ext}, we extend the
model to accommodate classical models of population growth and
to handle several features common to ecological
time-series, such as excess zeros and unexplained random variation. 
%Specifically, we describe methods for accommodating
%excess zeros and nuisance variables such as random observer effects.
Both frequentist and Bayesian methods of inference are discussed, and
code for fitting models is presented in the appendices.
%The Bayesian method
%is attractive in that it can accommodate prior information about
%detection parameters, which is available for many species, yet not
%collected as part of many existing monitoring programs.
In Section~\ref{sec:app}, we evaluate the performance of the model
extensions using a simulation study and by analyzing data from the
North American Breeding Bird Survey (BBS), one of
the most spatially and temporally extensive sets of count data on
vertebrate populations \citep{robbins_etal:1986}. The overarching aim
of the paper is to provide ecologists with flexible and accessible means of
addressing important questions related to the variation of abundance in
space and time. 
%believe these extensions will increase the utility of the model for
%application to existing datasets such the Breeding Bird Survey.
\begin{comment}
  The DM model allows for inference about spatial and temporal
dynamics in abundance while accounting for observation error
using only spatially and temporally replicated count data. Their model
is an extension of the closed-population $N$-mixture model
\citep{royle:2004biom}, which was designed specifically to
address the problem of modeling spatial variation in abundance in the
  presence of observation error, specifically, when detection
  probability is less than one. The DM extension relaxes the
assumption of population closure and includes explicit
parameters describing population change as a first-order Markovian
  process. [more details]

@@

Although the DM model was designed explicitly for these
purposes,
the original formulation of the model makes strict
distributional
  assumptions and does not acknowledge many of the sources of
variation inherent to existing ecological time-series data.  We  propose variations on these models that are likely to be more
realistic and useful in many cases. We demonstrate the
usefulness
  of these new models with simulated and real BBS data.

@@

Our work has highlighted the limitations of classical state-space
models as applied to ecological time series data, and we have
demonstrated how a class of open population $N$-mixture models
proposed by \citet{dail_madsen:2011} overcomes these
limitations. We extended their model in several important
ways to accomodate common features of ecological
datasets, including sparse counts, zero-inflation, environmental
stochasticity, and random variation in observation error.
%We made four important developments of the open population
%N-mixture model proposed by \citet{dail_madsen:2011} that make the
%model more applicable to long-term data such as is collected
%by the North American Breeding Bird Survey.
%First, we have
%reconciled the objectives of traditional state-space models
%with open population $N$-mixture models by illustrating how
%classical models of population dynamics can be embedded within
%the framework. Second, we have demonstrated methods of
%accounting for zero-inflation in the time-series. Third, we
%have illustrated how additional random effects such as
%observer-specific detection probabilities can be
%accommodated. Fourth, we presented a Bayesian analysis of the
%model, which, makes it possible to incorporate prior information when
%available, and in some cases, facilitates parameter estimation.
We also demonstrated how many of the objectives of conventional
state-space modeling can be accomplished within this expanded
framework. For example, one of the primary aims of our paper was to demonstrate how classical
population growth models can be embedded in the DM model.
Although we view this as an important connection to make between the two
different classes of state-space models, we believe that
%In some ways,
such population growth models are more phenomenological
than mechanistic, and this runs counter to the motivation for
the DM model. However, estimating
demographic parameters from count data is an ambitious goal, and
the required assumptions will not be valid in many cases,
especially when immigration and emigration occur. This was supported by the
unrealistic estimates of vital rates reported here and by the
lack of support for the mechanistic models relative to the population
growth models.

>>>>

\documentclass[12pt]{article}
\renewcommand*\abstractname{Summary}
\usepackage[a4paper,hmargin={2.4cm,2.4cm},vmargin={2.4cm,2.4cm}]{geometry}
\usepackage[mathlines]{lineno} %,displaymath
\usepackage{setspace}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{bm}
\usepackage{graphicx}
\usepackage[round,colon,authoryear]{natbib}
\usepackage{color, soul}
\usepackage{fancyvrb}
\usepackage{verbatim}
%\usepackage[pdftex,colorlinks=true,citecolor=black,linkcolor=black,urlcolor=black,pdftex,pdfstartview=FitH,bookmarks=true]{hyperref}
\usepackage[pdftex,hidelinks,pdfstartview={XYZ 1 1 1.5},bookmarks=true]{hyperref}
\usepackage{rotating}

@@

\abstract{
\noindent
\begin{enumerate}
\item Models of population dynamics %play a central role in theoretical and
%applied ecology, where they 
are frequently used for purposes such as testing 
hypotheses about density dependence and predicting species' responses 
to future environmental change or conservation actions. Fitting models
of population dynamics to field data is challenging because most data
sets are characterized by missing data and observation error, which
can inflate estimates of process variation if ignored.  
%Failure to account for observation error in such models can result in 
%bias. Therefore, ecologists have increasingly relied on 
Recently, state-space models have been developed to deal with these problems
%these probdirectly model 
by directly modeling both the observation error and the ecological
process of interest.
\item Conventional state-space models, however, have
several important limitations: (1) they assume that random effects are
Gaussian distributed, which implies that abundance can be negative and
that false negative observation errors are equally likely as false positive
errors, %results in unrealistic predictions
%(e.g. negative values of abundance) and an implausible mode
%and observation error are Gaussian random effects
%the parameters are not identifiable
%in many common situations, 
(2) they do not admit spatial variation in
population dynamics, and (3) some of the parameters of the model are not
estimable. % they assert an implausible model for
%we have no clear interpretation of the
%observation error. 
We demonstrate how each of these problems can be
resolved using a class of hierarchical models proposed by
\citet[Biometrics]{dail_madsen:2011} 
%for spatially-replicated time series data 
that attributes observation error to imperfect detection.
\item We expand this class of models to accommodate classical growth
models (e.g. exponential and Ricker-logistic), zero-inflation,
and random effects. 
% such as observer-specific detection probabilities. 
We also present methods for forecasting
population size under future environmental conditions. Implementation
of these ideas is possible using either frequentist or Bayesian
methods, as demonstrated by accompanying {\bf R} and {\bf JAGS} code.%. We provide code to fit these models using the \textbf{R} package
%\texttt{unmarked} and \textbf{JAGS} is also provided.
\item Results of a simulation study suggest that bias is %was 
negligible and coverage
nominal for the proposed model extensions. An analysis of data from
the North American Breeding Bird Survey highlights how these methods
can be readily applied to existing data, but it also suggests 
that precision will be low when direct information about detection
probability (such as is collected using distance sampling or
replicated counts) is lacking.
\end{enumerate}
}

@@

Recently developed state-space models %are a widely used approach for
have made it possible to %become the primary tool for 
study population dynamics while accounting for both process variation
and observation error \citep{devalpine_hastings:2002,
  buckland_etal:2004, dennis_etal:2006}. Classical state-space
models are time series models in which the true state of the
system (e.g. population size during each year) is modeled as a hidden
process, and the observed data are modeled conditional on this process
and the observation error. One reason for the widespread adoption of
state-space models in ecology is that failure to account for 
%process variation and 
observation error can bias estimators of abundance and
population growth parameters. For instance, the strength of
density dependence will be over-estimated if observation error is
ignored \citep{link_nichols:1994,shenk_etal:1998}.

A simple state-space model can be described as follows.
Let $N_t$ be the abundance of a species during year $t$, for
$t=1,\hdots,T$, and let $X_t$ be
the observed data, which differs from $N_t$ due to observation error,
a random effect denoted $\zeta_t$. Temporal variation in $N_t$ is
modeled using a population growth model, $\mu(N_{t-1})$
and random process variation denoted $\eta_t$.
% Richard, is switching this from $\grave{o}$ to $\eta$ okay?  Andy thought
% the o with accent was weird and I think I agree.  $\nu$ seems standard,
% but we use it for a slightly different meaning (ES only) later.  
The growth model may be density-dependent, as in the case of the 
Ricker-logistic model, or it might be density-independent, such as when growth
is exponential.
%, in which $\mu(N_{t-1}) = N_{t-1}e^{r}$ where $r$ is the intrinsic rate of increase.
A generic model can be written: %The full model can now be written as:
\begin{subequations}
  \label{eq:ss1}
  \begin{align}
    N_1 &= X_1 \label{eq:ss1a} \\
N_t &= \mu(N_{t-1}) + \eta_{t-1} \quad \text{for} \;
t=2,\hdots,T \label{eq:ss1b} \\
X_t &= N_t + \zeta_t \qquad \qquad \;\, \text{for} \;
t=1,\hdots,T. \label{eq:ss1c}
  \end{align}
\end{subequations}
%where $\eta_t$ is the random effect allowing for process
%variation unaccounted for by the deterministic model. 
In classical
state-space models, the two sets of random effects
are assumed to be i.i.d. Gaussian deviates:
$\eta_t \sim \mathrm{N}(0, \sigma)$ and
$\zeta_t \sim \mathrm{N}(0, \tau)$. %Standard practice is to
%ignore 
The process variation associated with $N_1$ is often ignored, or
%, as indicated by Eq~\ref{eq:ss1a}. Alternatively, 
the population is assumed to be at equilibrium such that the $N_1$ can
be regarded as an outcome of the equilibrium distribution.

Even though state-space models such as that shown in Eq.~\ref{eq:ss1}
overcome many of the limitations associated with the application of
classical models of population dynamics, 
%are among the most widespread approaches for modeling population dynamics
%using time series data, 
several problems exist. %, which we explain in the following paragraphs. 
% First,
% time series data and the underlying abundance parameters of interest
% are typically integer-valued, as in the case of count data, raising
% concerns about the Gaussian assumption for the random
% effects. Second, the model for observation error is implausible. %has little biological basis. 
% Third, the model does not account for spatial variation in abundance. Lastly, some of the
% parameters of the model are not estimable. We briefly discuss each of
% these points before describing a general approach to resolving these
% shortcomings.
%Use of the Gaussian distribution for modeling random process variation
%and observation error is often motivated by convenience rather than
%biology. Specifically, t
The assumption that random effects are Gaussian distributed makes it
easier to estimate parameters because methods such as the Kalman
filter can be applied %The Gaussian assumption allows for efficient methods of parameter
%estimation, such as the Kalman filter
\citep{dennis_etal:2006}; however, % which are not available otherwise % which is much more computationally efficient
%than estimation methods when random effects are not Gaussian distributed
%\citep{devalpine_hastings:2002}. %The problem with this is that it
%However, 
these models effectively assume that abundance can be take on any
real value, and hence they can predict negative values of abundance. %including negative values, 
%This approach allows for negative and non-integer values of the state-variable, which is
%inconsistent with the observed data, and may 
%and these models can result in implausible predictions.
Transforming the count data and the underlying abundance variables 
do little to resolve these issues \citep{ohara_kotze:2010}.

With respect to the observation error, the Gaussian assumption
%Another problem with standard state-space models is that the model for
%the mean zero Gaussian distribution for the observation error
implies that false positive detections are
equally likely as false negatives, a pattern inconsistent with most
findings \citep{miller_etal:2011}. 
%. To our knowledge, all studies to
%date have shown that false negative detections are much more common
%that false positives \citep{miller_etal:2011}.
%will be
%higher than $N_t$ as often as it is lower than $N_t$. 
%Identifying a biological mechanism that would cause such symmetric errors
%is difficult.
A more likely form of observation error, and one that has been recognized for well
over a century, results from failing to detect individuals that are
present. Imperfect detection may be attributable to
characteristics of the species under study, such as its elusiveness,
or to individuals being missed by the ecologist collecting the data in the field.
Although a vast number of methods have been devised for accounting for
this form of observation error, rarely have these methods been
integrated into state-space models of population dynamics \citep[but
see][]{buckland_etal:2004}.

Other limitations of state-space models, as commonly applied in ecology,
include the failure to admit spatial variation in abundance and
the fact that some parameters are not identifiable
\citep{polansky_etal:2009}. 
% without making 
%Another limitation of many state-space models is that they do 
%Common formulations of state space models do 
%not admit spatial variation in abundance or demographic
%parameters. 
Ignorning spatial variation simply reduces the scope of the
inferences, 
% that can 
%be drawn from the models, it disregards the fact that space is an
%important factor in the regulation of natural populations \citep{murdoch:1994,
%pulliam:1988, gill_etal:2001}.
but identifiability problems require %lack of identifiability because
model constraints or the assertion of additional assumptions. 
%A more serious problem %than the ones associated with the Gaussian assumptions 
%is that the parameters of simple state-space models
%such as Eq~\ref{eq:ss1} are not identifiable in many
%circumstances \citep{polansky_etal:2009}. 
Specifically, abundance in
the first year cannot be estimated because there is only one data
point (the count in the first year) that carries any information about
it. To remedy this, pratitioners either assume that there is no
observation error in the first year, or they assume the population is
at equilibrium, which may defeat the purpose of many studies of
population dynamics. 
%Recall that the
%Eq.~\ref{eq:ss1} did not specify distributions for $N_1$, % or $X_1$;
%although $N_1$ is a random variable. Thus, uncertainty exists
%that should be accounted for in the model.  
%%are needed to fully specify the model.
%To be more specific, a fully-specified state-space model requires at least
%three probability distributions, which we represent
%using bracket notation (i.e., $[Y|\Omega]$ is interpreted
%as the probability distribution of the random variable $Y$ given the
%parameter $\Omega$). The three probability distributions required for a
%state-space model are:
%%%% JEFF, does this extra sentence help? This notation comes from
%%%% several of the original state-space papers and is widespread in
%%%% books like Andy's and Bill's.
%\hl{[I'm not familiar with bracket notation,
% so these equations are confusing to me. How widely used is this
% notation in ecology? Will others likely understand this? Also, we might want
%to save space by trimming a bunch of this discussion back down.]}
%\begin{subequations}
%  \label{eq:ss2}
%  \begin{align}
%    [N_1&|\bm{\theta}] \label{eq:ss2a} \\ \label{eq:ss2b}
%[N_t&|N_{t-1},\bm{\Theta}] \quad \text{for} \; t=2,\hdots,T \\
%\label{eq:ss2c}
%    [X_t&|N_t,\bm{p}]  \qquad \; \text{for} \; t=1,\hdots,T
%  \end{align}
%\end{subequations}
%where $\bm{\theta}$ are the process variation parameters for the
%initial state, $\bm{\Theta}$ are the process variation parameters
%influencing how abundance changes over time, and $\bm{p}$ are the
%observation error parameters. As mentioned previously, Eq.~\ref{eq:ss2b}
%and Eq.~\ref{eq:ss2c} are assumed to be Gaussian in classical state-space
%models, but what should be the distribution for Eq.~\ref{eq:ss2a}?
%The three lines of general model
%described by Eq.~\ref{eq:ss2} correspond to the three equations shown
%for the specific example in Eq~\ref{eq:ss1}. However, in this case,
%we allow $N_1$ to be a random variable. But what should its
%distribution, $[N_1|\bm{\theta}]$ be?
%And how can $\bm{\theta}$ be estimated given that there is only a single observation
%vailable? 
%Often, these problems are simply assumed away. For
%example, sometimes researchers assume that there is no process variation in
%the first year, although this seems unjustified. Alternatively, the
%population may be assumed to be at equilibrium such that the expected
%value of $N_t$ is constant through time. Although this 
%makes the parameters $\bm{\theta}$ identifiable, assuming equilibrium
%defeats the objective of many studies of population dynamics, namely
%determining why a population varies over time.
%% OR
%Sometimes researchers assume that there is no process variation in 
%the first year or that the population is at equilibrium such that the expected
%value of $N_t$ is constant through time, but these approaches are unsatisfactory.

@@

In this paper, we focus on the model of \citet[henceforth the DM model]{dail_madsen:2011}
that resolves each of the problems outlined above. 
%simultaneous resolves each of the aforementioned problems with
%traditional state-space models, and is designed for simple count data.
%In the following section, we describe the DM model in its original
%form and explain how it resolves each of the deficiencies with
%standard state-space models. 
%In Section~\ref{sec:ext}, 
Our aim is to extend the
model to accommodate classical models of population growth and
to handle several features common to ecological
time series, such as excess zeros and unexplained random variation. 
%Specifically, we describe methods for accommodating
%excess zeros and nuisance variables such as random observer effects.
Both frequentist and Bayesian methods of inference are discussed, and
we evaluate the performance of the model using a simulation study by
analyzing data from the North American Breeding Bird Survey (BBS), one of
the most spatially and temporally extensive sets of count data on
vertebrate populations \citep{robbins_etal:1986}.
%code for fitting models is presented in the appendices.
%The Bayesian method
%is attractive in that it can accommodate prior information about
%detection parameters, which is available for many species, yet not
%collected as part of many existing monitoring programs.
%In Section~\ref{sec:app}, we evaluate the performance of the model
%extensions using a simulation study and by analyzing data from the
%North American Breeding Bird Survey (BBS), one of
%the most spatially and temporally extensive sets of count data on
%vertebrate populations \citep{robbins_etal:1986}. %The overarching aim
%of the paper is to provide ecologists with flexible and accessible means of
%addressing important questions related to the variation of abundance in
%space and time. 
%believe these extensions will increase the utility of the model for
%application to existing datasets such the Breeding Bird Survey.
\begin{comment}
  The DM model allows for inference about spatial and temporal
dynamics in abundance while accounting for observation error
using only spatially and temporally replicated count data. Their model
is an extension of the closed-population $N$-mixture model
\citep{royle:2004biom}, which was designed specifically to
address the problem of modeling spatial variation in abundance in the
  presence of observation error, specifically, when detection
  probability is less than one. The DM extension relaxes the
assumption of population closure and includes explicit
parameters describing population change as a first-order Markovian
  process. [more details]

@@

Although the DM model was designed explicitly for these
purposes,
the original formulation of the model makes strict
distributional
  assumptions and does not acknowledge many of the sources of
variation inherent to existing ecological time series data.  We  propose variations on these models that are likely to be more
realistic and useful in many cases. We demonstrate the
usefulness
  of these new models with simulated and real BBS data.

@@

Our work has highlighted the limitations of classical state-space
models as applied to ecological time series data, and we have
demonstrated how a class of open population $N$-mixture models
proposed by \citet{dail_madsen:2011} overcomes these
limitations. We extended their model in several important
ways to accomodate common features of ecological
datasets, including sparse counts, zero-inflation, environmental
stochasticity, and random variation in observation error.
%We made four important developments of the open population
%N-mixture model proposed by \citet{dail_madsen:2011} that make the
%model more applicable to long-term data such as is collected
%by the North American Breeding Bird Survey.
%First, we have
%reconciled the objectives of traditional state-space models
%with open population $N$-mixture models by illustrating how
%classical models of population dynamics can be embedded within
%the framework. Second, we have demonstrated methods of
%accounting for zero-inflation in the time series. Third, we
%have illustrated how additional random effects such as
%observer-specific detection probabilities can be
%accommodated. Fourth, we presented a Bayesian analysis of the
%model, which, makes it possible to incorporate prior information when
%available, and in some cases, facilitates parameter estimation.
We also demonstrated how many of the objectives of conventional
state-space modeling can be accomplished within this expanded
framework. For example, one of the primary aims of our paper was to demonstrate how classical
population growth models can be embedded in the DM model.
Although we view this as an important connection to make between the two
different classes of state-space models, we believe that
%In some ways,
such population growth models are more phenomenological
than mechanistic, and this runs counter to the motivation for
the DM model. However, estimating
demographic parameters from count data is an ambitious goal, and
the required assumptions will not be valid in many cases,
especially when immigration and emigration occur. This was supported by the
unrealistic estimates of vital rates reported here and by the
lack of support for the mechanistic models relative to the population
growth models.
